### 2012-08-31 Dail_Mad_ext.tex
methods

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}

@@

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

@@

\section{Methods}

\subsection{Model Development}

As originally formulated, the DM model includes submodels for four
conditionally related processes: the initial abundance state, apparent
survival, recruitment, and detection probability. Dail and Madsen
proposed two options for initial abundance (Poisson and negative
binomial):
(2)
(3)
where $N_{i1}$  is the abundance at site $i$  during year 1 and $\Lambda$ is
the expected abundance. The negative binomial model includes the
overdispersion parameter $\alpha$.  The survival and recruitment terms allow
abundance to change over time:
(4)
where $\omega$ is the survival probability and $\gamma$ is the arrival rate (which
can depend on Nit-1). Dail and Madsen propose three models for $\gamma$: the
constant model, where $\gamma$ does not depend on Nit-1, and which simulates
a ``propagule rain'' of recruitment; the autoregressive model , which
simulates geometric or density independent growth; and the ``no-trend''
model, which keeps abundance fixed over time.
The number seen is a function of detection probability:
(5)
Initial abundance, apparent survival, recruitment, and detection can
all be modeled as functions of covariates, allowing fairly
sophisticated inference about population processes and dynamics from
point count data.
Range model of initial abundance
Dail and Madsen (2011) suggested two distributions for modeling
initial abundance: Poisson and negative binomial.  We have extended
their model to include another distribution, the zero-inflated
Poisson.  This could be useful when, for example, one is modeling the
abundance of several bird species with the same set of BBS routespoint
count surveys for all species, but this set includes routes sites
outside the range of some of the bird species.  The distribution of
initial abundances can be represented as:
(6)
where $\psi$ represents the proportion of extra zeros.
These models allow three sources of zero counts by observers: a
species was at a route but not detected; the route was within the
species' range but there were no birds at that site in that year; and
the route was outside the species' range.  Furthermore, detection,
abundance, and zero-inflation can be modeled separately as functions
of different (or the same) covariates.  For example, detection of
species x at site y in year z might depend on wind speed, abundance on
forest type and weather, and zero-inflation upon elevation and
climate.  This approach combines elements of occupancy modeling
(MacKenzie et al. 2006) and abundance modeling, and could prove useful
for distinguishing species' fundamental and realized niches
(Hutchinson 1957).   The zero inflation factor can also affect

>>>>

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}

@@

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

@@

\section{Model Development}

\subsection{Original Formulation}

As originally formulated, the DM model includes submodels for four
conditionally related processes: the initial abundance state, apparent
survival, recruitment, and detection probability. \citet{dail_madsen:2011}
proposed two distributions for initial abundance, the Poisson and negative
binomial:
\begin{gather}
N_{it} \sim \mathrm{Pois}(\Lambda) \nonumber \\
N_{it} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1
and $\Lambda=\mathbb{E}[N_{i1}]$ is the expected abundance. The negative
binomial model includes the overdispersion parameter $\alpha$.  The
survival and recruitment terms allow abundance to change over time:
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the survival probability and $\gamma$ is the arrival
rate (which can depend on $N_{it-1}$). Dail and Madsen propose three models
for $\gamma$: the constant model, where $\gamma$ does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of recruitment; the
autoregressive model , which
simulates geometric or density independent growth; and the ``no-trend''
model, which keeps abundance fixed over time.

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  \label{eq:2}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
\end{equation}

Initial abundance, apparent survival, recruitment, and detection can
all be modeled as functions of covariates, allowing fairly
sophisticated inference about population processes and dynamics from
point count data.

\subsection{Range model of initial abundance}

\citet{dail_madsen:2011} suggested two distributions for modeling
initial abundance: Poisson and negative binomial.  We have extended
their model to include another distribution, the zero-inflated
Poisson.  This could be useful when, for example, one is modeling the
abundance of several bird species with the same set of BBS routespoint
count surveys for all species, but this set includes routes sites
outside the range of some of the bird species.  The distribution of
initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\Lambda &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.

These models allow three sources of zero counts by observers: a
species was at a route but not detected; the route was within the
species' range but there were no birds at that site in that year; and
the route was outside the species' range.  Furthermore, detection,
abundance, and zero-inflation can be modeled separately as functions
of different (or the same) covariates.  For example, detection of
species x at site y in year z might depend on wind speed, abundance on
forest type and weather, and zero-inflation upon elevation and
climate.  This approach combines elements of occupancy modeling
(MacKenzie et al. 2006) and abundance modeling, and could prove useful
for distinguishing species' fundamental and realized niches
(Hutchinson 1957).

The zero inflation factor can also affect
